Misexpression of germline programs with loss of the X-linked intellectual disability factor KDM5C

### 2024-04-18 submission/manuscript.Rmd
add EpiLC draft text and figure

@@

>>>>

## _Kdm5c_-KO cells express master regulators of germline identity in an _in vitro_ model of early embryogenesis

Misexpression of germline genes in the _Kdm5c_-KO brain suggests mutants fail to demarcate germline and somatic cellular identity. Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014] when a subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the body's somatic tissues[@bardotMouseGastrulationCoordination2020]. This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into post-implantation epiblast-like stem cells (EpiLCs) (Figure 2A). Previous studies have demonstrated that while some germline-enriched genes are also expressed in embryonic stem cells (ESCs), they are silenced as they differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021]. Therefore, to elucidate the mechanism underlying aberrant germline genes expression in the mature _Kdm5c_-KO brain, we first assessed their expression in male (XY) and female (XX) _Kdm5c_ mutant EpiLCs using our previously generated RNA-seq dataset[@samantaActivationXistEvolutionarily2022]. Because complete loss of KDM5C in females is not lethal at this time point, we were able to characterize germline transcripts in both _Kdm5c_ heterozygous (XX _Kdm5c_-HET) and homozygous null (XX _Kdm5c_-KO) female EpiLCs.

We found both XX and XY _Kdm5c_ mutant EpiLCs aberrantly express many germline transcripts in comparison to sex-matched wild-type controls (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1)(Figure 2B). While many germline genes were dysregulated in all _Kdm5c_ mutants, female heterozygous and homozygous mutants expressed about twice as many germline genes than _Kdm5c_-KO males (XX _Kdm5c_-HET: 212  XX _Kdm5c_-KO: 186, XY _Kdm5c_-KO: 90) (Figure 2C). We then wondered if the abundance of germline transcripts in _Kdm5c_ mutant females could be due to defects in X chromosome inactivation caused by KDM5C loss[@samantaActivationXistEvolutionarily2022], since the X chromosome is enriched with many spermatogenesis genes[@wangAbundanceXlinkedGenes2001]. However, the majority of germline DEGs for both males and females are autosomal genes (Figure 2C), indicating increased germline gene misexpression in females is independent of any potential X chromosome inactivation defects. We then asked if the identity of ectopic germline transcripts was influenced by _Kdm5c_ mutant sex, as some germline genes show biased transcription in the wild-type XX or XY gonads. We found XX _Kdm5c_-HET, XX _Kdm5c_-KO, and XY _Kdm5c_-KO EpiLCs all expressed similar proportions of XX-biased, XY-biased, and unbiased germline genes that roughly reflected the proportion of genes shared and unique within the male and female wild-type gonads (In WT gonads: 6% XX-biased, 79% XY-biased, 15% unbiased) (Figure 2D). *Note: test if unbiased genes are over-represented, I think % wise they are*

To evaluate whether germline mRNAs are constitutively dysregulated or change over the course of development, we compared EpiLC germline DEGs to those expressed in the _Kdm5c_ mutant brain. The majority of EpiLC germline DEGs were unique to EpiLCs, with only a few shared across brain sequencing datasets (Figure 2E). Only two germline genes were consistently misexpressed in all male and female datasets, _DNA segment, Chr 1, Pasteur Institute 1 (D1Pas1)_ and _Testis expressed gene 14 (Tex14)_. Surprisingly, germline DEGs unique to _Kdm5c_ mutant EpiLCs included master regulators of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 2F). These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but they are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019]. _Stra8_'s expression is typically sexually dimorphic, as female germ cells initiate meiosis during embryogenesis while males begin meiosis after birth [@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019]. Yet we found both XX and XY _Kdm5c_ mutant EpiLCs ectopically express _Stra8_ RNA. XX and XY _Kdm5c_ mutants also aberrantly express _Dazl_, a translational regulator essential for germ cell function[@liDAZLMasterTranslational2019]. Interestingly, a significant portion of germline transcripts expressed in _Kdm5c_ mutant EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (XY _Kdm5c_-KO EpiLCs p = 1.698e-07, Fisher’s Exact Test). This suggests expression of DAZL protein would enable the translation of other aberrant germline transcripts, influencing their ability to impact on _Kdm5c_ mutant cellular function. Therefore, we then stained XY _Kdm5c_-KO EpiLCs for DAZL protein expression through immunocytochemistry (Figure 2G). We found about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein in their cytoplasm (p = 0.0015, Welch's t-test, XY KO versus WT), consistent with the subcellular localization observed when DAZL is stained in spermatogonia[@mikedisDAZLMediatesBroad2020].

@@

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/EpiLCs_too.pdf}
  \caption[Male and female \textit{Kdm5c} mutant epiblast-like cells express master regulators of germline identity ]{\textbf{Male and female \textit{Kdm5c} mutant epiblast-like cells express master regulators of germline identity A.} Top - Diagram of \textit{in vivo} differentiation of emrbyonic stem cells (ESCs) of the inner cell mass into epiblast stem cells and primordial germ cells. Bottom - \textit{in vitro} differentiation of ESCs into epiblast-like cells (EpiLCs). \textbf{B.} Left - Bigwigs of an example germline gene, \textit{Cyct} that is dysregulated in male and female \textit{Kdm5c} mutant EpiLCs. Right - Overlap of all germline DEGs between male and female EpiLCs. \textbf{C.} Chromosome location of germline genes misexpressed in male and female EpiLCs. \textbf{D.} Number of EpiLC germline DEGs that show sex-biased expression in the wild-type gonads. \textbf{E.} Upset plot displaying the overlap of germline DEGs expressed in male and female \textit{Kdm5c} mutant EpiLC and brain RNA-seq datasets. \textbf{F.} Expression of genes key for early germ cell development in \textit{Kdm5c} mutant EpiLCs in transcrips per million (TPM) with p-values for Welch's t-test compared to wild type of the same sex. \textbf{G.} Immunocytochemistry of DAZL in male wild-type (WT) and \textit{Kdm5c}-KO (5CKO) EpiLCs.}
  \label{figurelabel}
\end{figure}
